Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-09-17 paper/manuscript.tex
Includes updates to software packages and tweaks to the setup of hydration free energies from @IAlibay, responses to reviewer comments from @apbhati, and updates to the author lists.

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[18]{Stefan Boresch}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[11, 12, 13]{Peter V. Coveney}
\author[15]{Jonathan W. Essex}
\author[19]{Emilio Gallicchio}
\author[9]{Vytautas Gapsys}
\author[23,24]{Michael Gillhofer}
\author[9]{David F. Hahn}
\author[20]{Wei-Tse Hsu}
\author[1,10]{Maximilian Kuhn}
\author[15]{Oliver J. Melling}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[23,24]{Chris Oostenbrink}
\author[4]{Meghan Osato}
\author[15,16]{William G. Poole}
\author[6]{Samarjeet Prasad}
\author[21,22]{Benjamin Ries}
\author[2,7]{Andrea Rizzi}
\author[17]{Jenke Scheen}
\author[14]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
\affil[15]{School of Chemistry and Chemical Engineering, University of Southampton, Southampton, SO17 1BJ, United Kingdom}
\affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
\affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
\affil[18]{University of Vienna, Faculty of Chemistry, Institute of Computational Biological Chemistry, Wahringer Str. 17, A-1090 Vienna}
\affil[19]{Department of Chemistry and Biochemistry, Brooklyn College of the City University of New York, New York, NY, 11210
Ph.D. Program in Chemistry, The Graduate Center of the City University of New York, New York, NY, 10016
Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}
\affil[20]{Department of Biochemistry, University of Oxford, Oxford OX1 3QU, United Kingdom}
\affil[21]{Boehringer Ingelheim Pharma GmbH \& Co KG, Medicinal Chemistry, Birkendorfer Str 65, 88397 Biberach an der Riss, Germany}
\affil[22]{Open Free Energy, Open Molecular Software Foundation, Davis, CA, 95616, United State}
\affil[23]{Institute of Molecular Modeling and Simulation, BOKU University, 1190 Vienna, Austria}
\affil[24]{Christian Doppler Laboratory for Molecular Informatics in the Biosciences, BOKU University, 1190 Vienna, Austria}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Stefan Boresch}{0000-0002-2793-6656}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Emilio Gallicchio}{0000-0002-2606-4913}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}
\orcid{William G. Poole}{0009-0003-2441-8794}
\orcid{Oliver J. Melling}{0000-0001-6243-7433}
\orcid{Jonathan W. Essex}{0000-0003-2639-2746}
\orcid{Wei-Tse Hsu}{0000-0001-6167-5480}
\orcid{Benjamin Ries}{0000-0002-0945-8304}
\orcid{Meghan Osato}{0000-0002-2732-457X}
\orcid{Michael Gillhofer}{0000-0001-8081-6373}
\orcid{Chris Oostenbrink}{0000-0002-4232-2556}

@@

\paragraph{The choice of solvent force field impacts the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. In a study by Open Force Field, they found that the GAFF\cite{wang2004gaff} and Sage \cite{boothroyd2023sage} forcefields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as RESP\cite{bayly1993resp} or AM1-BCC\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points. This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points. The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points. The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, TIES~\cite{bhati2017} circumvents this issue by design as it does not need any restraints.

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017}.

@@

Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be "ergodic" but it must also be ``mixing''~\cite{coveney2016, MDbook2025}. Rate processes and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

 The precision of NEQ calculations is determined by the overlap between the forward and reverse work distributions corresponding to out-of-equilibrium transitions between the two end-points. To estimate the overlap, Wu \& Kofke suggest using dissipated work values which can be shown to correspond to the relative entropies of the densities in the phase space. \cite{wu2005phase} Another approach to assess NEQ convergence relies on calculating the first two moments of the Fermi function in Bennett Acceptance Ratio estimator. \cite{hahn2010measuring} Since the second order estimate converges slower than the first, the overall convergence of the free energy estimate can be quantified as the normalized difference between the two moments of the Fermi function. In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work. \cite{shirts2005comparison} Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well. They recommended an alternative metric, "distance" defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.

@@

	\item \href{https://www.charmm.org/}{CHARMM} has a variety of tools developed over the years. The PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module.  

@@

	\item \href{https://biosimspace.org/}{BioSimSpace} is a multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace}. 
	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM. 

@@

@@

@@

	\item \href{https://github.com/MobleyLab/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.

@@

@@

@@

@@

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[22]{Irfan Alibay}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[18]{Stefan Boresch}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[26]{Finlay Clark}
\author[11, 12, 13]{Peter V. Coveney}
\author[15]{Jonathan W. Essex}
\author[19]{Emilio Gallicchio}
\author[9]{Vytautas Gapsys}
\author[23,24]{Michael Gillhofer}
\author[9]{David F. Hahn}
\author[20]{Wei-Tse Hsu}
\author[25]{Sheenam Khuttan}
\author[1,10]{Maximilian Kuhn}
\author[15]{Oliver J. Melling}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[23,24]{Chris Oostenbrink}
\author[4]{Meghan Osato}
\author[15,16]{William G. Poole}
\author[6]{Samarjeet Prasad}
\author[21,22]{Benjamin Ries}
\author[2,7]{Andrea Rizzi}
\author[17]{Jenke Scheen}
\author[14]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
\affil[15]{School of Chemistry and Chemical Engineering, University of Southampton, Southampton, SO17 1BJ, United Kingdom}
\affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
\affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
\affil[18]{University of Vienna, Faculty of Chemistry, Institute of Computational Biological Chemistry, Wahringer Str. 17, A-1090 Vienna}
\affil[19]{Department of Chemistry and Biochemistry, Brooklyn College of the City University of New York, New York, NY, 11210
Ph.D. Program in Chemistry, The Graduate Center of the City University of New York, New York, NY, 10016
Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}
\affil[20]{Department of Biochemistry, University of Oxford, Oxford OX1 3QU, United Kingdom}
\affil[21]{Boehringer Ingelheim Pharma GmbH \& Co KG, Medicinal Chemistry, Birkendorfer Str 65, 88397 Biberach an der Riss, Germany}
\affil[22]{Open Free Energy, Open Molecular Software Foundation, Davis, CA, 95616, United State}
\affil[23]{Institute of Molecular Modeling and Simulation, BOKU University, 1190 Vienna, Austria}
\affil[24]{Christian Doppler Laboratory for Molecular Informatics in the Biosciences, BOKU University, 1190 Vienna, Austria}
\affil[25]{SandboxAQ, Palo Alto, California 94301, United States}
\affil[26]{School of Natural and Environmental Sciences, Newcastle University, Newcastle upon Tyne NE1 7RU, United Kingdom}

@@

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Stefan Boresch}{0000-0002-2793-6656}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Emilio Gallicchio}{0000-0002-2606-4913}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}
\orcid{William G. Poole}{0009-0003-2441-8794}
\orcid{Oliver J. Melling}{0000-0001-6243-7433}
\orcid{Jonathan W. Essex}{0000-0003-2639-2746}
\orcid{Wei-Tse Hsu}{0000-0001-6167-5480}
\orcid{Benjamin Ries}{0000-0002-0945-8304}
\orcid{Meghan Osato}{0000-0002-2732-457X}
\orcid{Michael Gillhofer}{0000-0001-8081-6373}
\orcid{Chris Oostenbrink}{0000-0002-4232-2556}
\orcid{Irfan Alibay}{0000-0001-5787-9130}
\orcid{Sheenam Khuttan}{0000-0002-8873-6359}
\orcid{Finlay Clark}{0000-0003-0474-5475}

@@

\paragraph{The choice of solvent force field impacts the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation of the system ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. In a study by Open Force Field, they found that the general AMBER force field (GAFF)\cite{wang2004gaff} and Open Force Field Sage \cite{boothroyd2023sage} force fields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as Restrained Electrostatic Potential (RESP)\cite{bayly1993resp} or AM1 population charges with bond-charge correction (AM1-BCC)\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. It involves a unique post-processing of the initial MCSS to ensure that the final MCSS is both structurally and chemically identical and has a single set of coordinates while maintaining dual topology for the alchemical parts.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

\paragraph{Ring breaking and forming.} \label{subsec: Ring breaking and forming}
Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES overcomes this issue by ensuring that MCSS atoms are structurally and chemically identical while allowing for partial rings~\cite{bhati2017, bieniek2021}.

@@

Statistical mechanics provides a route to calculate macroscopic thermodynamic properties of a system from \textit{ensemble averaging} over microstates. Hitherto, MD practitioners have been assuming the validity of the ergodic hypothesis that allows one to equate ensemble averaging to time averaging in the limit of time approaching infinity to predict thermodynamic properties using one-off simulations. However, it is often ignored that this limit is on the order of a Poincaré recurrence time, which is truly astronomical and well beyond the reach of today's supercomputers. In more rigorous terms, ergodic theory necessitates that, if a dynamical system is to approach equilibrium, it is not sufficient for it to be ``ergodic'' but it must also be ``mixing'' (these terms refer to the different classes of dynamical systems provided by ergodic theory; please see the references for details)~\cite{coveney2016, MDbook2025}. Rate processes (a course of change within a system that occurs over time) and equilibration times are a property of the distribution function, not of individual trajectories. Consequently, neighbouring trajectories in the underlying phase space diverge exponentially which means that a single trajectory can never capture the true behaviour of a system. Rather, one must perform true ensemble averaging through running sufficiently large number of replicas, termed ``ensemble simulations''. This is manifest in the numerous published reports where authors have highlighted the non-reproducibility of results obtained from one-off simulations and that single simulations heavily underestimate the true uncertainty in predictions~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. The uncertainty is calculated as the standard error of the mean for estimates obtained from each replica in the ensemble. The ensemble size is a parameter that needs to be determined empirically for the system of interest and may need to be adjusted; however, some general recommendations (for both EQ and NEQ approaches) based on systematic analyses on a large and diverse dataset of protein-ligand complexes are available in a series of papers published by the Coveney group~\cite{bhati2017, bhati2018, bhati2019, bhati2022, wan2023eqvsneq, bhati2025}. Further, ensemble simulations are also necessary in case of enhanced sampling methods~\cite{bhati2018} as well as microsecond long simulations~\cite{bhati2023}. It is the influence of these reports that journals such as \textit{JCIM} and \textit{Communications Biology} have now mandated the use of ensemble simulations in computational studies~\cite{wan2023}.

@@

 The precision of NEQ calculations is determined by the overlap between the forward and reverse work distributions corresponding to out-of-equilibrium transitions between the two end-points. To estimate the overlap, Wu \& Kofke suggest using dissipated work values which can be shown to correspond to the relative entropies of the densities in the phase space. \cite{wu2005phase} Another approach to assess NEQ convergence relies on calculating the first two moments of the Fermi function in Bennett Acceptance Ratio estimator. \cite{hahn2010measuring} Since the second order estimate converges slower than the first, the overall convergence of the free energy estimate can be quantified as the normalized difference between the two moments of the Fermi function. In practice, analytical error estimate of the BAR estimator \cite{shirts2003equilibrium} gives an indication when the overlap is lacking: the estimated error rapidly grows with the increase in the dissipated work. \cite{shirts2005comparison} Bhati \textit{et al.} recently studied the effect of such overlaps on the accuracy of predicted RBFEs~\cite{wan2023eqvsneq} as well as ABFEs~\cite{bhati2025} but found that overlap (or lack of it) does not capture the reliability of calculations very well. They recommended an alternative metric, ``distance'' defined as the difference between the inner extremes of the two work value distributions, which correlates quite well with the unsigned errors in free energy predictions and hence was found a better indicator of the reliability of NEQ calculations.

@@

	\item \href{https://www.charmm.org/}{CHARMM} has a variety of tools developed over the years. The PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyse the data after the MD run. Lambda-dynamics-based free energy calculation can be carried out using the BLOCK module. GPU optimized implementation of CHARMM's lambda dynamics is also implemented in the new package BLaDE \cite{hayes2021blade}. A number of energy interpolation schemes have been implemented in the new GPU package apoCHARMM \cite{prasad2025apocharmm}.     

@@

	\item \href{https://biosimspace.org/}{BioSimSpace} is a multiscale molecular simulation framework, written to allow computational modellers to quickly prototype and develop new algorithms for molecular simulation and molecular design~\cite{hedges2019biosimspace, hedges2023suite}. 
	\item \href{https://siremol.org/}{Sire} is a multiscale, molecular simulation framework that provides several applications, including SOMD, an MD/MC code for performing FEP calculations via an interface to OpenMM~\cite{woods2024sire}. 

@@

        \item \href{https://github.com/choderalab/openmmtools} provides tooling for creating alchemical simulations using OpenMM. These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints.

@@

    \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies. Current offerings, notably include a Perses-based \url{https://github.com/choderalab/perses/tree/main/perses} hybrid topology method using OpenMM.

@@

	\item \href{https://github.com/OpenFreeEnergy/Lomap}{Lomap/Lomap2} : Relative alchemical transformation graph planning for setting up perturbation networks~\cite{liu2013lead}.
        \item \href{https://github.com/OpenFreeEnergy/kartograf} 
    Kartograf: An geometry-based atom mapper for setting up relative free energies with a conserved core~\cite{ries2024kartograf}.
        \item \href{https://github.com/OpenFreeEnergy/Konnektor}
    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{}.

@@

        \item 
    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.

@@

\textbf{IA} Reviewed and contributed to Sec. ~\ref{sec:prerequisites}, ~\ref{subsec:hydration}, and ~\ref{sec:multiple-chemical-states}. Updated Sec. ~\ref{sec:software}.

@@

\textbf{FC}: Coordinated the v2.0 updates to the article (TODO: be more specific?/ mention ABFE updates?). \\

@@

\textbf{SK}: Contributed to Sec. \ref{sec:atm}.\\
